---
title: "COSMOS results of SI and Colon samples"
author: "A. Gabor"
date: "11/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Here we present the results of the integrated omics analysis using COSMOS. 
COSMOS integrates transcriptomics and metabolomics data using prior knowledge. 
This prior knowledge is a network, where nodes represent metabolites and proteins
(e.g. kinases and metabolic enzymes) and signed, directed edges represent interactions
among these entities. COSMOS builds on an optimization framework, that finds a subset 
of the original prior knowledge network that is in agreement with the data.

We estimated transcription factor activity from the transcriptomics data using 
Dorothea regulon and Viper algorithm. The TF activity served as input nodes for COSMOS.
Then we defined the measured metabolites that changed significantly as outputs.

To explore multiple possible solutions for each samples, we (1) set in each COSMOS run to report 
multiple solutions within the given time limit (15 hours) and (2) we rerun the optimization 10 times
with shuffled inputs.

Here we summarise the obtained results. 




```{r message=FALSE, warning=FALSE, include=FALSE}
#library(cosmos)
library(here)
library(tidyverse)
source("./parse_CPLEX_log.R")
source("./format_cosmos_results_new.R")


clean_run = FALSE
```


```{r include=FALSE}
# The following samples were used for data integration because there the significant
# number of metabolites were more than 3. 

keep_sample  = c("colon_1000uM_24h", "colon_1000uM_48h",
				 "colon_1000uM_72h", "colon_100uM_24h",
				 "si_1000uM_24h",    "si_1000uM_48h",
				 "si_1000uM_72h")
samples = tibble(sample = keep_sample, sample_id = as.character(1:length(keep_sample)))

# read in metabolites names: 
# from the experiments: 
metabolomics <- read_rds("./data/carnival_inputs/human_metabolomics_data.rds")
metabolite_names = tibble(name = metabolomics$`possible annotation based on accurate mass`,
						  pubchem = metabolomics$pubchemID) %>% unique()

# metabolite names for the PKN:
metab_map <- read_csv(here("./data/metab_to_pubchem.csv")) %>%
	mutate(pubchem = pubchem)

# use the names in the experiments if there is a mapping in both datasets
# there is not a 1:1 mappng between name and puchem. Take the first name for each pubhchem id
metabolite_names = metab_map %>% 
	group_by(pubchem) %>% slice(1) %>%  # takes first name for each id. 
	filter(! pubchem %in% metabolite_names$pubchem ) %>%
	bind_rows(metabolite_names)
	

tf_data_list <- read_rds(here("data/carnival_inputs_cluster/tf_data.rds"))
tf_regulon <- read_rds(here("data/carnival_inputs_cluster/tf_regulon.rds"))
metabolomics_data_list <- read_rds(here("data/carnival_inputs_cluster/metabolomics_data.rds"))

# import cosmos results and parse the log files:
if(clean_run){
	results_data <- tibble(res_files = list.files("./data/cosmos_results",pattern = "[0-9]+.rds",full.names = TRUE)) %>%
		# load results
		mutate(cosmos_out = map(res_files,read_rds)) %>%
		# determine sample_id and repetition from file names
		mutate(sample_id = map_chr(res_files,function(fname){
			pname <- basename(fname) %>% gsub(".rds","",.) %>% strsplit(.,split = c("_"),fixed = TRUE) 
			pname[[1]][[1]]
		})) %>%
		mutate(repetition_id = map_chr(res_files,function(fname){
			pname <- basename(fname) %>% gsub(".rds","",.) %>% strsplit(.,split = c("_"),fixed = TRUE) 
			ifelse(length(pname[[1]]) == 1, "0", pname[[1]][[2]])
		})) %>%
		# construct the name of the corresponding log file
		mutate(log_file = 
			   	paste0("./data/carnival_logs/",
			   		   "carnival_",sample_id,
			   		   ifelse(repetition_id==0,"",paste0("_",repetition_id)),
			   		   ".out"
			   	)) %>%
		mutate(log = map(log_file,parse_CPLEX_log))
	
	results_data <- results_data %>% 
		left_join(samples,by="sample_id") %>%
		select(sample,cosmos_out,log,repetition_id,everything())
	
	write_rds(results_data,here("./data/analysis_data/results_data.rds"))
}else{
	results_data <- read_rds(here("./data/analysis_data/results_data.rds"))
	
}
```

## Quality of solutions
For each sample we run the COSMOS optimization 10 times, for 15 hours each, 
however in some cases the optimizaiton reached a memory limit of our cluster (128 GB),
which resulted in loss of some results.

The figure below shows that for each sample, three to eight runs finished. 

```{r include=FALSE}
# arrange the optimisation results in a single line
tidy_optim_results <- function(log_item){
	tibble(GAP = last(log_item$convergence$Gap),
		   `GAP [%]` = last(log_item$convergence$`Gap [%]`),
		   objective = last(log_item$objective),
		   n_solution = last(log_item$n_solutions),
		   termination_reason = last(log_item$termination_reason))
}


optim_results <- results_data %>% select(sample,repetition_id,log) %>%
	mutate(optim_stats = map(log,tidy_optim_results)) %>%
	unnest(optim_stats)
```


```{r echo=FALSE}
optim_results %>% ggplot() + geom_bar(aes(sample)) + 
	ggtitle("Number of finished optimisation runs") + 
	theme_bw() + theme(axis.text.x = element_text(angle=45,hjust = 1))
```
In each run we instructed the optimizer to enumerate multiple solutions, however, 
in most cases it found only a few (6-9) solutions. There are two samples, where
many equivalent solutins were found. 

```{r echo=FALSE}
optim_results %>% group_by(sample) %>%
	summarise(total_solutions = sum(n_solution),.groups="drop") %>%
	ggplot() + geom_col(aes(sample,total_solutions)) +
	geom_text(aes(sample,total_solutions,label=total_solutions),nudge_y = 10) +
	ggtitle("Number of found solutions") + 
	theme_bw() + theme(axis.text.x = element_text(angle=45,hjust = 1))
```


### Objective functions:
The objective function measures the quality of each optimization. The objective function has two parts 

(1) the sum of absolute difference between the model simulation and measurements and
(2) the number of edges used to explain the data. 

therefore a smaller value is better. 

```{r echo=FALSE}
optim_results %>% ggplot() + 
	geom_jitter(aes(sample,objective,col=repetition_id),width = 0.1,height = 0) + 
	ggtitle("Final objective function values") + 
	theme_bw() + 
	theme(axis.text.x = element_text(angle=45,hjust = 1))
```

This figure shows that the repeated runs achieved similar objective functions. This give 
us confidence that they found a global optimal solution. 

The samples _colon_1000uM_24h_ and _colon_100uM_24h_ have high objective function
which means that the number of measurements are not well explained in those cases. 
We check the reason below.

Since the objective functions are similar in each sample, we can aggregate the multiple runs. 



## Explained inputs and measurements
How many inputs and outputs are explained by the inferred networks?

```{r include=FALSE}
# merge the cosmos inputs and outputs:
tf_data = enframe(tf_data_list,name = "sample",value="tf_data")
metabolomics_data = enframe(metabolomics_data_list, name = "sample",value="metabolomics_data")

cosmos_total <- results_data %>%
	left_join(tf_data,by = "sample")%>% 
	left_join(metabolomics_data,by = "sample")
```

```{r echo=FALSE}
explained_variables <- cosmos_total %>% 
	mutate(explained_nodes = pmap(.,function(cosmos_out,tf_data,metabolomics_data,...){
		
		tibble(n_tfs = length(tf_data),
			   modelled_tfs = sum(names(tf_data) %in% c(cosmos_out$aggregated_network$source,cosmos_out$aggregated_network$target)),
			   modelled_tfs_rel = modelled_tfs/length(tf_data),
			   n_metabolites = length(metabolomics_data),
			   modelled_metabolites = sum(names(metabolomics_data) %in% c(cosmos_out$aggregated_network$source,cosmos_out$aggregated_network$target)),
			   modelled_metabolites_rel = modelled_metabolites/length(metabolomics_data))
		
	})) %>% unnest(explained_nodes) %>%
	group_by(sample) %>%
	summarise(across(.cols = c(9:14),mean))
	

explained_variables %>% select(sample,n_tfs,modelled_tfs,n_metabolites,modelled_metabolites) %>% print()
```
The above table shows how many of the transcription factors (`n_tfs`) and metabolites (`n_metabolites`)
are explained in average by the COSMOS results (`modelled_tfs` and `modelled_metabolites`). 

There are a few reasons why not all of them are included: 

- there is a trade-off between model complexity (number of edges) and fitness: each edge in the model 
adds to the objectve function, therefore it is better to leave out some nodes than included many edges. 
- early termination of the optimization: optimization might needed more time to integrate more nodes, but
this also means higher memory usage, which was limited. 
- possible controversiality between data and prior knowledge. 


The above numbers explain why samples `colon_1000uM_24` and `colon_100uM_24` had high objective
function value. These samples had 400+ metabolites as inputs but only 30% of them 
are included in the solutions. 


```{r message=FALSE, warning=FALSE, include=FALSE}
# Rename the proteins and metabolites to user friendly names: 
omnipath_ptm = read_rds("./data/carnival_inputs/omnipath_ptm.rds")

metabolite_names_vec = metabolite_names$name
names(metabolite_names_vec) = metabolite_names$pubchem

clean_metabolic_expression <- function(metab_name){
	
	metab_name %>% gsub("^Metab__","",x = .) %>%
		gsub("___([a-z])____","_\\1",x = .) 
	
}

clean_enzyme_expression <- function(enzyme_name){
	
	enzyme_name %>% gsub("^Enzyme[0-9]+__","",x = .) %>%
		gsub("_reverse","",x = .) %>%
		gsub("EXCHANGE[1-9]+","",x = .) 
	
}

extract_clean_names <- function(node_table){
	

	
	node_table %>% mutate(clean_names = Nodes) %>%
		# fix metabolite names: removes compartments
		mutate(clean_names = ifelse(grepl("^Metab",clean_names),clean_metabolic_expression(clean_names),clean_names)) %>%
		mutate(clean_names = ifelse(grepl("^Enzyme",clean_names),clean_enzyme_expression(clean_names),clean_names)) %>%
		left_join(mutate(metab_map,pubchem = as.character(pubchem)),by = c("clean_names" = "pubchem")) %>%
		mutate(clean_names = ifelse(!is.na(name),name,clean_names)) %>%
		select(-name) %>%
		dplyr::select(Nodes,clean_names,everything()) %>%
		mutate(compartment = ifelse(type == "metabolite",gsub(".*_","",clean_names),NA)) %>%
		mutate(clean_names = ifelse(type == "metabolite",gsub("_[a-z]","",clean_names),clean_names))
}



cosmos_networks <-  cosmos_total %>%
	dplyr::select(-log,-res_files,-sample_id,-log_file) %>%
	dplyr::mutate(network_raw = pmap(.,function(cosmos_out,tf_data,metabolomics_data,...){
		
		res <- format_COSMOS_res(cosmos_res = cosmos_out,
								 metab_mapping = metabolite_names_vec,
								 gene_mapping = "org.Hs.eg.db",
								 measured_nodes = c(names(metabolomics_data),names(tf_data)),
								 omnipath_ptm = omnipath_ptm
		)
		
		edges = res[[1]]
		nodes = res[[2]]
		nodes_updated <- extract_clean_names(nodes)
		return(list(edges,nodes_updated))
		
	}))
```




```{r include=FALSE}
show_network <- function(graph,title){
	
	edges= graph[[1]] %>% dplyr::rename(from = source,
										to = target) %>%
		
		mutate(color = "black") %>%
		mutate(value = weight) %>%
		mutate(arrows.to.type = c("circle","arrow")[as.factor(interaction)])
	
	nodes = graph[[2]] %>% as_tibble() %>%
		filter(Nodes %in% c(edges$from, edges$to)) %>% 
		dplyr::rename(id=Nodes) %>%
		# work on the graph format
		dplyr::mutate(label = clean_names) %>%
		#dplyr::slice(148) %>%
		dplyr::mutate(shape = purrr::map_chr(type, function(x){
			switch(x, "metabolite" = "diamond",
				   "metab_enzyme" ="square",
				   "TF" = "star", 
				   "protein" ="ellipse",
				   "Kinase" = "ellipse")	
			
		} )) %>%
		dplyr::mutate(borderWidth = ifelse(measured,1,0)) %>%
		mutate(color.background = c("red","blue")[as.factor(Activity)],
			   color.border = c("white","black")[as.factor(measured)])
	
	
	vn <- visNetwork::visNetwork(nodes = nodes,edges = edges,main = title)  %>%
		visNetwork::visNodes() %>%
		visNetwork::visEdges(smooth = FALSE) %>%
		visNetwork::visPhysics(stabilization = FALSE) %>%
		visNetwork::visIgraphLayout()
	return(vn)
	
}
cosmos_networks <- cosmos_networks %>% mutate(visnet_network = map2(network_raw,paste(sample,repetition_id),show_network))
# cosmos_networks$visnet_network[[1]]
```


## Aggregated network solutions
Since the objective function values of the optimized models are similar, we 
can merge them. In the visualization we weight each edge by the time it 
appears across the solutions. This weight shows "essentiality" of the edges, i.e. 
the edges that disappears in some solutions are not essential to explain the data. 

Note, if an edge appears in all solutions, it does not necessarily mean that there 
is no alternative explanation. It can happen that the optimizer didn't find this 
alternative solution or the alternative solution includes more edges -- our approch
seeks for the smallest possible model. 


```{r include=FALSE}
aggregate_interactions <- function(network,N_sol){
	
	sifs = lapply(network,function(x)x[[1]])
	names(sifs) = paste0("network",1:length(sifs))
	
	sif_all <- bind_rows(sifs,.id = "network")
	solutions <- tibble(N_sol = N_sol, network =  paste0("network",1:length(sifs)))
	
	## edge_weigth * number of solutions  gives the number of times the edge appeared. 
	# then appearance can be sum-ed across repetitions
	sif_all <- sif_all %>%
		left_join(solutions,by="network") %>%
		mutate(appearance = N_sol * weight) %>%
		group_by(source,interaction,target) %>%
		summarise(total_appearance = sum(appearance),.groups = "drop" ) %>%
		mutate(total_rel_appearance = total_appearance/sum(N_sol))
	
	return(list(sif_all))
}

aggregate_nodes <- function(network,N_sol){
	
	nodes = lapply(network,function(x)x[[2]])
	names(nodes) = paste0("network",1:length(nodes))
	
	nodes_all <- bind_rows(nodes,.id = "network")
	solutions <- tibble(N_sol = N_sol, network =  paste0("network",1:length(nodes)))
	
	## AvgAct * number of solutions  gives the number of times the edge was 
	# activated-inactivated. 
	# then appearance can be sum-ed across repetitions
	nodes_all <- nodes_all %>% 
		left_join(solutions,by="network") %>%
		mutate(net_activity = N_sol * AvgAct) %>%
		group_by(Nodes,clean_names,NodeType,measured,type) %>%
		summarise(total_activity = sum(net_activity),.groups = "drop" ) %>%
		mutate(total_rel_activity = total_activity/sum(N_sol))
	
	return(list(nodes_all))
}

consensus_cosmos_networks <- cosmos_networks %>% dplyr::select(sample, cosmos_out,network_raw) %>%
	mutate(N_networks = map_dbl(cosmos_out,function(x)x$N_networks)) %>%
	group_by(sample) %>%
	dplyr::summarise(consensus_nodes = aggregate_nodes(network_raw,N_networks),
					 consensus_edges = aggregate_interactions(network_raw,N_networks),.groups="drop")


# remove those nodes that are not in the edge table

consensus_cosmos_networks <- consensus_cosmos_networks %>% mutate(consensus_nodes  = map2(consensus_nodes,consensus_edges,function(nodes,edges){
	
	 filtered_nodes <- nodes %>% filter(Nodes %in% c(edges$source,edges$target))
	
}))


```




```{r include=FALSE}

show_consensus_network <- function(interactions, nodes,title){
	
	edges= interactions %>% dplyr::rename(from = source,
										  to = target) %>%
		
		mutate(color = "black") %>%
		mutate(value = total_rel_appearance) %>%
		mutate(arrows.to.type = c("circle","arrow")[as.factor(interaction)])
	
	nodes = nodes %>% as_tibble() %>%
		filter(Nodes %in% c(edges$from, edges$to)) %>% 
		dplyr::rename(id=Nodes) %>%
		# work on the graph format
		dplyr::mutate(label = clean_names) %>%
		# simplify labels:
		
		#dplyr::slice(148) %>%
		dplyr::mutate(shape = purrr::map_chr(type, function(x){
			switch(x, "metabolite" = "diamond",
				   "metab_enzyme" ="square",
				   "TF" = "star", 
				   "protein" ="ellipse",
				   "Kinase" = "ellipse")	
			
		} )) %>%
		dplyr::mutate(borderWidth = ifelse(measured,1,0)) %>%
		mutate(color.background = c("red","blue")[as.factor(sign(total_rel_activity))],
			   color.border = c("white","black")[as.factor(measured)])
	
	
	N = visNetwork::visNetwork(nodes = nodes,edges = edges,main = title)  %>%
		visNetwork::visNodes() %>%
		visNetwork::visEdges(smooth = FALSE) %>%
		visNetwork::visPhysics(stabilization = FALSE) %>%
		visNetwork::visIgraphLayout()
	
	return(N)
}

networks <- pmap(consensus_cosmos_networks,~with(list(...),show_consensus_network(consensus_edges,consensus_nodes,sample)))

```


The following figures show the consensus networks of each sample.
Please note that you can zoom in to the network and rearrange the nodes if needed. 

Legend: 

Node type is encoded by the shape:

- Metabolites - diamond
- Metabolic enzyme - square
- transcription factor - star
- other proteins - ellipse

Colors:

- down-regulation - red 
- up-regulation - blue

Black border represents input and output nodes: estimated trascription factors and measured metabolites. 


```{r}
networks[[1]]
```


```{r}
networks[[2]]
```


```{r}
networks[[3]]
```


```{r}
networks[[4]]
```


```{r}
networks[[5]]
```


```{r}
networks[[6]]
```

```{r}
networks[[7]]
```

### Export sample networks to cytoscape
```{r}
cytoscape_folder = "./data/cosmos_samples_cytoscape_inputs"
dir.create(cytoscape_folder,showWarnings = FALSE)
consensus_cosmos_networks	%>% mutate(pmap(.,function(sample,consensus_edges,consensus_nodes,...){
		sif_file = paste0(cytoscape_folder, "/",sample,"_sif.tsv")
		attribute_file = paste0(cytoscape_folder, "/",sample,".tsv")
		edges = consensus_edges
		nodes = consensus_nodes
		edges$interaction = ifelse(edges$interaction==1,"up","down")
		
		edges <- edges %>% rename(essentiality = "total_rel_appearance") %>%
			select(-total_appearance)
		nodes <- nodes %>% rename(name = "clean_names",activity="total_rel_activity") %>%
			select(-NodeType,-total_activity)
		
		write_tsv(select(edges,source,interaction,target,essentiality),file = sif_file )
		write_tsv(nodes,file = attribute_file )
		return(TRUE)
		
}))
```






### edge similarities
We clustered the samples based on the edges in their networks:
rows represent the edges, samples are in the columns. The color shows the essentiality of the edge. 

Early time SI samples are the most similar and late time colon samples also cluster together. 


```{r echo=FALSE, fig.width=3, fig.height=4}
edge_table <- consensus_cosmos_networks %>% select(sample,consensus_edges) %>%
	unnest(consensus_edges) %>%
	mutate(source = ifelse(grepl("^Metab",source),clean_metabolic_expression(source),source)) %>%
	mutate(source = ifelse(grepl("^Enzyme",source),clean_enzyme_expression(source),source)) %>%
	mutate(target = ifelse(grepl("^Metab",target),clean_metabolic_expression(target),target)) %>%
	mutate(target = ifelse(grepl("^Enzyme",target),clean_enzyme_expression(target),target)) %>%
	mutate(edge = paste(source,target,sep = " -> ")) %>%
	select(sample,edge,total_rel_appearance) %>%
	group_by(sample,edge) %>%
	summarise(exist = max(total_rel_appearance),.groups = "drop") 


edge_table %>%
	spread(sample,exist,fill = 0) %>% column_to_rownames("edge") %>%
	pheatmap::pheatmap(show_rownames = FALSE)
```

We filter for edges that appear across multiple samples to look for common mechanism.
The following heatmap shows the edges for the samples that appears in at least 4 samples.  

There are some interactions that appear exclusively for colon samples and some that
appear across all the samples. 

```{r, fig.height=12,fig.width=7, echo=FALSE}
common_edges <- edge_table %>% group_by(edge) %>%
	summarise(n_samples = n(),.groups="drop") %>% 
	arrange(desc(n_samples)) %>%
	filter(n_samples > 3) %>% pull(edge)


edge_table %>% filter(edge %in%common_edges ) %>%
	spread(sample,exist,fill = 0) %>% column_to_rownames("edge") %>%
	pheatmap::pheatmap(show_rownames = TRUE)
```

### Node similarities
We clustered the samples based on the nodes in their networks:
rows represent the nodes, samples are in the columns.

Early time SI samples are the most similar and late time colon samples also cluster together. 


```{r echo=FALSE, fig.height=4, fig.width=3}
node_table <- consensus_cosmos_networks %>% select(sample,consensus_nodes) %>%
	unnest(consensus_nodes) %>%
	select(sample,clean_names,total_rel_activity) %>%
	group_by(sample,clean_names) %>%
	summarise(exist = median(total_rel_activity),.groups = "drop") 


node_table %>%
	spread(sample,exist,fill = 0) %>% column_to_rownames("clean_names") %>%
	pheatmap::pheatmap(show_rownames = FALSE)
```

We filter for nodes that appear across multiple samples to look for common mechanism.
The following heatmap shows the nodes for the samples that appears in at least 3 samples.  



```{r, fig.height=15, fig.width=5, echo=FALSE}
common_nodes <- node_table %>% group_by(clean_names) %>%
	summarise(n_samples = n(),.groups="drop") %>% 
	arrange(desc(n_samples)) %>%
	filter(n_samples > 3) %>% pull(clean_names)


node_table %>% filter(clean_names %in% common_nodes ) %>%
	spread(sample,exist,fill = 0) %>% column_to_rownames("clean_names") %>%
	pheatmap::pheatmap(show_rownames = TRUE,main = "Node activity")
```



## Pathways in the solution networks
Which genes-sets are represented in results more than random? We apply over-representation
analysis to answer this question. 
We use MSigDB gene sets: Hallmark genes and Canonical pathways for the analysis. 

**Hallmark gene sets** summarize and represent specific well-defined biological
states or processes and display coherent expression.

**Canonical pathways:** Gene sets from pathway databases. Usually, these gene
sets are canonical representations of a biological process compiled by
 domain experts.

```{r include=FALSE}
# ConsensusPathDB-human:
#  formatting to piano compatible data.frame
# CPDB_genesets <- read_tsv("./data/CPDB_pathways_genes.tab")
# CPDB_genesets_df <- CPDB_genesets %>% 
# 	filter(source == "Reactome") %>%
# 	select(pathway,hgnc_symbol_ids ) %>% 
# 	mutate(gene_list = strsplit(hgnc_symbol_ids,split = ",")) %>%
# 	select(-hgnc_symbol_ids) %>%
# 	unnest(gene_list)
# CPDB_gsc = piano::loadGSC(CPDB_genesets_df[,2:1])

msigdb <- read_rds("./data/msigdb_Genesets_Dec19.rds")

msig_to_df <- function(msig_collection){
	enframe(msig_collection) %>% unnest(value) %>%
		rename(geneset = name, gene=value) %>%
		select(gene,geneset)
}


# background genes: all the genes that appear in the PKN
meta_network = cosmos::load_meta_pkn()
prots <- unique(c(meta_network$source,meta_network$target))
prots <- prots[!grepl("XMetab",prots)]
prots <- gsub("^X","",prots)
prots <- gsub("Gene[0-9]+__","",prots)
prots <- gsub("_reverse","",prots)
prots <- gsub("EXCHANGE.*","",prots)
prots <- unique(prots)
prots <- unlist(sapply(prots, function(x){unlist(strsplit(x,split = "_"))}))

entrezid <- prots
gene_mapping <- AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, entrezid, 'SYMBOL', 'ENTREZID')
gene_mapping <- unlist(gene_mapping)
gene_mapping <- gene_mapping[!is.na(gene_mapping)]
background_genes <- unique(gene_mapping)

# over_rep_analysis <- consensus_cosmos_networks %>% mutate(CPDB_pws = map(consensus_nodes,function(nodes){
# 	
# 	genes_in_solutions = nodes$clean_names[nodes$clean_names %in% background_genes] %>%unique()
# 	
# 	CPDB_ora <- piano::runGSAhyper(genes = genes_in_solutions,
# 					   universe = background_genes,
# 					   gsc = CPDB_gsc)
# 	
# 	ora_results <- CPDB_ora$resTab %>% as_tibble() %>% arrange(`Adjusted p-value`) %>%
# 		mutate(pathway = rownames(CPDB_ora$resTab)) %>%
# 		select(pathway,everything())
# 	
# })) 

# runs overrepresentation analysis for given genes and genesets
run_ora <- function(nodes,edges,background,piano_geneset_obj){
	
	
	genes_in_solutions <- nodes %>% filter(Nodes %in% c(edges$source,edges$target))%>%
		filter(clean_names %in% background) %>%
		pull(clean_names) %>% unique()
	#genes_in_solutions = nodes$clean_names[nodes$clean_names %in% background] %>% unique()
	
	ora <- piano::runGSAhyper(genes = genes_in_solutions,
					   universe = background,
					   gsc = piano_geneset_obj)
	ora_results <- ora$resTab %>% as_tibble() %>% arrange(`Adjusted p-value`) %>%
		mutate(pathway = rownames(ora$resTab)) %>%
		select(pathway,everything())
}


hallmark_set <- msig_to_df(msigdb$MSIGDB_HMARKS) %>% 
	filter(gene %in%background_genes ) %>%
	piano::loadGSC() 

ora_hallmarks <- consensus_cosmos_networks %>%
	mutate(hallmark_pws = map2(.x = consensus_nodes,.y = consensus_edges,.f = run_ora,background = background_genes,piano_geneset_obj = hallmark_set)) %>%
	unnest(hallmark_pws) %>% select(-consensus_nodes,-consensus_edges)


canonical_set <- msig_to_df(msigdb$MSIGDB_CANONICAL) %>% 
	filter(gene %in%background_genes ) %>%
	piano::loadGSC() 

ora_canonical <- consensus_cosmos_networks %>%
	mutate(canonical_pws = map2(.x = consensus_nodes,.y = consensus_edges, .f = run_ora,background = background_genes,piano_geneset_obj = canonical_set)) %>%
	unnest(canonical_pws) %>% select(-consensus_nodes,-consensus_edges)


```



```{r,fig.width=20, fig.height= 15}
ora_hallmarks %>%
	group_by(sample) %>% slice_min(n=25,order_by = `Adjusted p-value`) %>%
	ggplot() + geom_col(aes(reorder(pathway,-`Adjusted p-value`),-log10(`Adjusted p-value`))) +
	geom_hline(yintercept = -log10(0.05),color="red",linetype = 2) +
	facet_wrap(~sample,scales = "free_y") + coord_flip() +theme_bw() + ggtitle("ORA using Hallmarks")
```





```{r,fig.width=20, fig.height= 15}
ora_canonical %>%
	group_by(sample) %>% slice_min(n=25,order_by = `Adjusted p-value`) %>%
	ggplot() + geom_col(aes(reorder(pathway,-`Adjusted p-value`),-log10(`Adjusted p-value`))) +
	geom_hline(yintercept = -log10(0.05),color="red",linetype = 2) +
	facet_wrap(~sample,scales = "free_y") + coord_flip() +theme_bw() + 
	ggtitle("ORA using Canonical Pathways")
```


### Pathway: TNFA_SIGNALING_VIA_NFKB

```{r}

HALLMARK_TNFA_SIGNALING_VIA_NFKB_genes <- msig_to_df(msigdb$MSIGDB_HMARKS) %>%
	filter(geneset=="HALLMARK_TNFA_SIGNALING_VIA_NFKB")

# is there some genes that appears across all the samples? 

consensus_cosmos_networks %>% select(sample,consensus_nodes) %>%
	unnest(consensus_nodes) %>%
	filter(clean_names %in% HALLMARK_TNFA_SIGNALING_VIA_NFKB_genes$gene) %>%
	group_by(sample) %>% 
	summarise(genes_in_set = paste(clean_names,collapse = ", "),.groups="drop")
```

## Filtering network for relevant pathways

pathways found in paper:

- Cell Cycle (Cell Cycle)
- DNA replication (DNA Replication)
- DNA synthesis (Synthesis of DNA)
- DNA damage (?)
- p53 signaling
- Apoptosis
- Respiratory electron transport and ATP synthesis
- Metabolism




```{r}
pw_file <- here("./data/List relevant pathways and DEGs.xlsx")
sheets <- readxl::excel_sheets(pw_file)

pw_table_raw <- tibble(sheet_name =sheets ) %>% 
	mutate(sheet = map(sheet_name,~readxl::read_excel(path = pw_file,sheet = .x))) 


pw_table <- pw_table_raw %>% mutate(sample = gsub(" ","uM_",sheet_name)) %>%
	select(sample,sheet) %>%
	mutate(sheet = map(sheet, ~gather(.x, key = pathway, value = genes))) %>%
	unnest(sheet) %>%
	filter(!is.na(genes)) %>%
	mutate(gene_symb = AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'SYMBOL', 'ENSEMBL'))

					
```


```{r}
# pw_networks: contains a network for each gene-set and each sample. 
pw_networks <- pw_table %>% 
	nest(pw_genes =c(genes,gene_symb) ) %>%
	full_join(consensus_cosmos_networks, by="sample") %>%   ## Full-join! 
	filter(!is.na(pathway))

```

# Find transcription factors: 

```{r}

# example
# sample = pw_networks$sample[[8]]
# pathway = pw_networks$pathway[[8]]
# consensus_nodes = pw_networks$consensus_nodes[[8]]
# consensus_edges = pw_networks$consensus_edges[[8]]
# pw_genes = pw_networks$pw_genes[[8]]

# Overlap of network nodes and genesets


pw_overlap <- pw_networks %>% 
	mutate(pmap_dfr(.,function(sample,pathway,pw_genes,consensus_nodes,consensus_edges){
	
	# check which node are in the pathway
	nodes_in_patway <- consensus_nodes %>%
		mutate(in_pathway = clean_names %in% pw_genes$gene_symb) %>%
		filter(in_pathway==TRUE) 
	
	
	return(tibble(nodes_in_patway = list(nodes_in_patway)))
	
})) %>% select(sample,pathway,nodes_in_patway) %>%
	unnest(nodes_in_patway) %>%
	select(sample,pathway,Nodes) %>%
	group_by(sample,pathway) %>%
	summarise(n_nodes_shared  = n())
	



pw_networks_results <- pw_networks %>% 
	mutate(pmap_df(.,function(sample,pathway,pw_genes,consensus_nodes,consensus_edges){
	
	# check which node are in the pathway
	nodes_in_patway <- consensus_nodes %>%
		mutate(in_pathway = clean_names %in% pw_genes$gene_symb) %>%
		filter(in_pathway==TRUE) %>% pull(Nodes)
	
	tfs_in_patway <- consensus_nodes %>%
		mutate(in_pathway = clean_names %in% pw_genes$gene_symb) %>%
		filter(in_pathway==TRUE, type=="TF") %>% pull(Nodes)
	
	edges = consensus_edges %>% dplyr::rename(from = source,
											  to = target)
	
	if(length(tfs_in_patway)==0) return(tibble(edges=list(NULL),nodes = list(NULL)))
	
	# look for the neighbourhood of these nodes and collect them
	G = consensus_edges %>% select(source,target,everything()) %>%
		igraph::graph_from_data_frame(.,directed = TRUE) 
	
	#ig_net <- igraph::make_ego_graph(G, nodes = tfs_in_patway, order = 3, mode = "out")
	
	# we check all nodes 
	ig_net <- igraph::make_ego_graph(G, nodes = c(tfs_in_patway,nodes_in_patway), order = 2, mode = "all")
	
	to_keep <- unlist(sapply(ig_net,function(x){igraph::V(x)$name}))
	
	
	edges = consensus_edges %>% dplyr::rename(from = source,
										  to = target) %>%
		filter(to %in% to_keep,from %in% to_keep) %>%
		mutate(essentiality = total_rel_appearance) 
	
	nodes = consensus_nodes %>%
		filter(Nodes %in% c(edges$from, edges$to)) %>% 
		dplyr::rename(id=Nodes) %>%
		# work on the graph format
		dplyr::mutate(label = clean_names) %>%
		mutate(in_pathway = id %in% nodes_in_patway)
	
	return(tibble(edges=list(edges),nodes = list(nodes)))
	
})) %>%
	# detect empty and remove them: 
	mutate(remove = map_lgl(edges,~is.null(.x))) %>%
	filter(!remove) %>% select(-remove)

edges <- pw_networks_results$edges[[1]]
nodes <- pw_networks_results$nodes[[1]]
pahway <- pw_networks_results$pathway[[1]]

pw_networks_visNet <- pw_networks_results %>%
	mutate(pmap_df(.,function(sample,pathway,edges,nodes,...){
		
		edges = edges %>%
			mutate(color = "black") %>%
			mutate(value = essentiality) %>%
			mutate(arrows.to.type = c("circle","arrow")[as.factor(interaction)]) %>%
			mutate(color = "grey")
		
		
		nodes = nodes %>%
			# work on the graph format
			dplyr::mutate(label = clean_names) %>%
			dplyr::mutate(shape = purrr::map_chr(type, function(x){
				switch(x, "metabolite" = "diamond",
					   "metab_enzyme" ="square",
					   "TF" = "triangle", 
					   "protein" ="dot",
					   "Kinase" = "dot")	
			} )) %>%
			dplyr::mutate(borderWidth = ifelse(in_pathway,3,0)) %>%
			mutate(color.background = c("#E41A1C","#999999","#377EB8")[factor(sign(total_rel_activity),levels = c(-1,0,1))],
				   color.border = c("white","black")[as.factor(in_pathway)]) %>%
			#mutate(color.background = ifelse(in_pathway,"green",color.background)) %>%
			mutate(size = ifelse(type=="TF",50,25))
		
		title = paste0(pathway,": ",sample)
		print(title)
		
		N = visNetwork::visNetwork(nodes = nodes,edges = edges,main = title)  %>%
			visNetwork::visNodes(font = list(size=20)) %>%
			visNetwork::visEdges(smooth = FALSE) %>%
			visNetwork::visPhysics(stabilization = FALSE) 		%>%
			#visNetwork::visHierarchicalLayout()
			visNetwork::visIgraphLayout()	
		
		return(tibble(edges=list(edges),nodes = list(nodes), vis_net = list(N)))
		
	}))




```




# Cell Cycle
```{r}
pw_networks_visNet %>% filter(pathway == "Cell cycle") %>%
	pull(vis_net)
```


```{r}
pw_networks_visNet %>% filter(pathway == "Metabolism") %>%
	pull(vis_net)
```

```{r}
pw_networks_visNet %>% filter(pathway == "Apoptosis") %>%
	pull(vis_net)
```



```{r}
pw_networks_visNet %>% filter(pathway == "DNA damage response") %>%
	pull(vis_net)
```




### Export to cytoscape sif and attribute fiels


```{r}

cytoscape_folder = "./data/cosmos_pathway_cytoscape_inputs"
dir.create(cytoscape_folder,showWarnings = FALSE)
pw_networks_results	%>% mutate(pmap(.,function(sample,pathway,edges,nodes,...){
		sif_file = paste0(cytoscape_folder, "/",paste0(pathway,"_",sample),"_sif.tsv")
		attribute_file = paste0(cytoscape_folder, "/",paste0(pathway,"_",sample),".tsv")
		
		edges$interaction = ifelse(edges$interaction==1,"up","down")
		
		edges <- edges %>% rename(source = "from",target="to")
		nodes <- nodes %>% rename(name = "clean_names",activity="total_rel_activity") %>%
			select(-NodeType,-total_activity)
		
		write_tsv(select(edges,source,interaction,target,essentiality),file = sif_file )
		write_tsv(nodes,file = attribute_file )
		return(TRUE)
		
}))
```



#### Export metabolic names

```{r}
network_metabs <- consensus_cosmos_networks$consensus_nodes[[1]] %>% filter(type=="metabolite")

#network_metabs %>% left_join(metabolomics,by = c(Nodes = "cosmosID"))
network_metabs %>% select(clean_names) %>% write_csv("./temp.csv")
```










